### 2017-11-09 manuscript.rmd
created manuscript_kernel.R

@@

	# Name of the compara gene trees file. This was downloaded from 
	# ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/
	# and compressed with gz

	gene_tree_name = "Compara.75.protein.nhx.emf.gz"

	# Files produced by KMRR Rscript.R with Tau values and other expression statistics
	expression_file_names = c(
		"ChickenBrawandTScomparisonTable_9_6chPC.txt", 
		"ChimpBrawandTScomparisonTable_9_6cmPC.txt", 
		"GorillaBrawandTScomparisonTable_9_6gPC.txt", 
		"HumBrawandTScomparisonTable_9_6hPC.txt", 
		"MacacaBrawandTScomparisonTable_9_6mcPC.txt", 
		"MusBrawandTScomparisonTable_9_6mPC.txt", 
		"OpossumBrawandTScomparisonTable_9_6oPC.txt", 
		"PlatypusBrawandTScomparisonTable_9_6pPC.txt"
	)

	kmrr_directory = "kmrr/"
	expression_file_names %<>% paste( kmrr_directory, ., sep="" )
	gene_tree_name %<>% paste( kmrr_directory, ., sep="" )

	# The minimum number of genes with expression data that a gene needs to have to be considered
	min_genes_with_expression = 4

	# These clade dates are from the KMRR code, they got them from http://timetree.org/
	calibration_times = data.frame(
		age =
			c( 20, 92, 167, 9, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
				414, 371, 162, 25, 74, 77, 86, 7 ), 
		clade = 
			c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
				"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
				"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
				"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
				"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires", "Hominini"
		),
		stringsAsFactors=FALSE
	)

	# The calibration times used for speciation nodes
	focal_calibrations_clades = 
		c( "Hominini", "Homininae", "Catarrhini", "Euarchontoglires", "Theria", "Mammalia", "Amniota" )

	# In simulation, the fold change in rate following duplication relative to speciation
	dup_adjust = 2

```{r define_functions, echo=FALSE, message=F, warning=F}

@@

```{r load_trees, echo=FALSE, cache=TRUE, message=F, warning=F}

	# Read in the tree file as a vector of strings, one per line
	lines = readLines( gene_tree_name )

	# Isolate strings that are trees
	# These start with (
	tree_lines = lines[ grepl( "^\\(", lines, perl = TRUE ) ]

	# Parse each string to a treeio::treedata object, adding a label column 
	# for node names
	gene_trees = mclapply( tree_lines, parse_gene_trees, mc.cores=cores )

	# Uncomment for development to look at a subset of gene trees
	# gene_trees = gene_trees[1:100]

	n_gene_trees = length( gene_trees )

	# Parse the tree annotations, these are lines that start with SEQ
	# These have the species names, which are not in the trees themselves
	tip_annotations = 
		lines[ grepl( "^SEQ", lines, perl = TRUE ) ] %>%	# Get lines that start with SEQ
	  str_replace( " \\(\\d+ of \\d+\\)", "" ) %>%      # Remove instances of '(1 of 2)' and '(2 of 2)'
	  str_replace( "(([^ ]+ ){8})([^ ]+) ([^ ]+)", "\\1\\3_\\4" ) %>%      # Get rid of spaces in last field of some entries
		str_c( collapse="\n" ) %>%												# Combine them into one string
		read_delim( delim=" ", col_names = FALSE ) %>% 		# Read string as tibble
		select( 2, 8 ) %>%																# Select the columns with species and gene ID
		rename( Ensembl.Gene.ID = X8 ) %>%
		rename( species = X2 )

	# Remove the large string objects to free up memory
	rm( lines )
	rm( tree_lines )

```{r load_expression, echo=FALSE, message=F, warning=F}

	# Read the expression data for each species and combine into a single tibble,
	# adding the species names from tip_annotations
	expression = 
		lapply(
			expression_file_names, 
			read.table, 
			stringsAsFactors=FALSE, 
			header=TRUE
		) %>%
		bind_rows() %>%
		left_join( tip_annotations, by="Ensembl.Gene.ID" )

	# Free up memory
	rm( tip_annotations )

```{r add_expression_to_trees, echo=FALSE, cache=TRUE, message=F, warning=F}

	# Annotate each tree by joining the corresponding expression data to the 
	# @data object. This maps the expression data to the tree tips
	gene_trees_annotated = mclapply(
		gene_trees,
		function( tree ){
			tree@data %<>% left_join( expression, by = c( "G" = "Ensembl.Gene.ID" ) )

			return( tree )
		}, 
		mc.cores=cores
	)

	# Free up memory
	rm( gene_trees )

	# Remove tips from the trees that do not have expression data
	gene_trees_pruned = mclapply( 
		gene_trees_annotated, 
		drop_empty_tips, 
		min_genes_with_expression=min_genes_with_expression,
		mc.cores=cores 
	)

	# Free up memory
	rm( gene_trees_annotated )

	# Remove trees with less than min_genes_with_expression tips with expression data
	gene_trees_pruned = gene_trees_pruned[ ! is.na( gene_trees_pruned ) ]

	n_removed_for_no_speciation = length( gene_trees_pruned )

	# Remove trees with no speciation events
	gene_trees_pruned = gene_trees_pruned[ unlist( lapply( gene_trees_pruned, get_n_speciation ) ) > 0 ]

	n_removed_for_no_speciation = n_removed_for_no_speciation - length( gene_trees_pruned )

	# fix clade names
	gene_trees_pruned = lapply( gene_trees_pruned, fix_hominini )

```{r calibrate_trees, echo=FALSE, cache=TRUE, message=F, warning=F}

	# Time calibrate all speciation nodes to the same years
	gene_trees_calibrated = calibrate_trees( gene_trees_pruned, calibration_times, model="correlated" )

```{r estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
	# Estimate model parameters and add them to the tree objects
	# This takes a while, so do it once here and reuse the values as needed
	gene_trees_calibrated = mclapply( 
		gene_trees_calibrated, 
		add_model_parameters, 
		mc.cores=cores 
	)
```

```{r trees_to_contrasts, cache=TRUE, echo=FALSE, message=F, warning=F}

	# Calculate the contrasts from the trees and associated expression data
	gene_trees_pic = add_pics_to_trees( gene_trees_calibrated )

	# Collect all the contrasts in a single tibble
	nodes_contrast = summarize_contrasts( gene_trees_pic )

	# Collect all the tree statistics in a single tibble
	tree_summary = summarize_trees( gene_trees_pic )

	# Test if the mean rank of duplication pics is greater than speciation pics
	wilcox_test_result = wilcox_oc( nodes_contrast )

@@

```{r ascertainment, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

	# Examine by node age

	max_speciation_age = 
		nodes_contrast %>% 
		filter( D=="N" ) %>%
		select( node_age ) %>%
		max( na.rm=TRUE )

	nodes_contrast_filtered_age = 
		nodes_contrast %>% 
		filter( node_age <= max_speciation_age )

	wilcox_filtered_age = wilcox_oc( nodes_contrast_filtered_age )

	# Examine by expected variance, ie branch length
	var_exp_max = 
		nodes_contrast %>% 
		filter( D=="N" ) %>%
		select( var_exp ) %>%
		max( na.rm=TRUE )

	var_exp_min = 
		nodes_contrast %>% 
		filter( D=="N" ) %>%
		select( var_exp ) %>%
		min( na.rm=TRUE )

	nodes_contrast_filtered_var = 
		nodes_contrast %>% 
		filter ( ( var_exp >= var_exp_min ) & ( var_exp <= var_exp_max ) )

	wilcox_filtered_var = wilcox_oc( nodes_contrast_filtered_var )

@@

```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}

	# Build a tibble of all pairwise comparisons between tips
	pairwise_summary = 
		mclapply( gene_trees_pic, get_pairwise_summary, mc.cores=cores ) %>% 
		bind_rows()

	# Calculate some summary statistics 
	ortholog_r = 
		pairwise_summary %>% 
		filter( D=="N" ) %$% 
		cor.test( tau_a, tau_b )

	paralog_r = 
		pairwise_summary %>% 
		filter( D=="Y" ) %$% 
		cor.test( tau_a, tau_b )

	paralog_subset_r = 
		pairwise_summary %>% 
		filter( D=="Y" ) %>% 
		filter( node_age <= max_speciation_age ) %$% 
		cor.test( tau_a, tau_b )

	ortholog_distance_mean = 
		pairwise_summary %>% 
		filter( D=="N" ) %>% 
		.[["distance"]] %>% 
		mean()

	paralog_distance_mean = 
		pairwise_summary %>% 
		filter( D=="Y" ) %>% 
		.[["distance"]] %>% 
		mean()

@@

```{r simulations_null, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

	# Replace the actual Tau values with simulated Tau values, using the parameters 
	# estimated for each tree
	gene_trees_sim_null = mclapply( gene_trees_calibrated, sim_tau, mc.cores=cores )

	# Perform contrasts 
	gene_trees_sim_null %<>% add_pics_to_trees()
	nodes_sim_null_contrast = gene_trees_sim_null %>% summarize_contrasts()

	# Calculate pairwise comparisons
	pairwise_sim_null_summary = 
		mclapply( gene_trees_sim_null, get_pairwise_summary, mc.cores=cores ) %>% 
		bind_rows()

	# Calculate some summary statistics 
	ortholog_r_sim_null = 
		pairwise_sim_null_summary %>% 
		filter( D=="N" ) %$% 
		cor.test( tau_a, tau_b )

	paralog_r_sim_null = 
		pairwise_sim_null_summary %>% 
		filter( D=="Y" ) %$% 
		cor.test( tau_a, tau_b )

```{r simulations_oc, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

	# Replace the actual Tau values with simulated Tau values
	gene_trees_sim_oc = lapply( gene_trees_calibrated, sim_tau, dup_adjust=dup_adjust )

	# Calculate contrasts 
	gene_trees_sim_oc %<>% add_pics_to_trees()
	nodes_sim_oc_contrast = gene_trees_sim_oc %>% summarize_contrasts()

	# Calculate pairwise comparisons 
	pairwise_sim_oc_summary = 
		mclapply( gene_trees_sim_oc, get_pairwise_summary, mc.cores=cores ) %>% 
		bind_rows()

	ortholog_r_sim_oc = 
		pairwise_sim_oc_summary %>% 
		filter( D=="N" ) %$% 
		cor.test( tau_a, tau_b )

	paralog_r_sim_oc = 
		pairwise_sim_oc_summary %>% 
		filter( D=="Y" ) %$% 
		cor.test( tau_a, tau_b )

@@

All files needed to re-execute the analyses presented in this document are available at https://github.com/caseywdunn/comparative_expression_2017. The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.

@@

```{r phylogenetic_signal, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

	k_thresh = quantile( tree_summary$K, na.rm=TRUE )[3] # third quantile is the median
	k_percentile = names(k_thresh)
	genes_pass_k = tree_summary$gene[ tree_summary$K > k_thresh ]

	nodes_contrast_filtered_k = 
		nodes_contrast %>% 
			filter( gene %in% genes_pass_k )

	wilcox_test_result_k = wilcox_oc( nodes_contrast_filtered_k )

@@

```{r model_selection, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

	dAIC_thresh = 2
	AIC_min = pmin( tree_summary$aic_bm, tree_summary$aic_ou )
	tree_summary$dAIC = tree_summary$aic_ou - tree_summary$aic_bm
	tree_summary$dAIC_bm = tree_summary$aic_bm - AIC_min
	tree_summary$dAIC_ou = tree_summary$aic_ou - AIC_min
	genes_pass_bm = tree_summary$gene[ tree_summary$dAIC_bm < dAIC_thresh ]
	genes_pass_ou = tree_summary$gene[ tree_summary$dAIC_ou < dAIC_thresh ]

	nodes_contrast_filtered_bm = 
		nodes_contrast %>% 
			filter( gene %in% genes_pass_bm )

	wilcox_test_result_bm = wilcox_oc( nodes_contrast_filtered_bm )

@@

```{r model_ou, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

	# Get the trees that pass AIC criteria for OU
	gene_trees_pic_digests = lapply( gene_trees_pic, digest )
	gene_trees_pic_ou = gene_trees_pic[ gene_trees_pic_digests %in% genes_pass_ou ]

	# Replace the BM pics with OU pics
	gene_trees_pic_ou %<>% add_pics_to_trees( model_method="OU" )
	nodes_ou_contrast = gene_trees_pic_ou %>% summarize_contrasts()
	wilcox_ou = wilcox_oc( nodes_ou_contrast )

@@

```{r calibration_sensitivity, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

	noised_replicate_n = 10
	sd_fraction = 0.2

	calibration_times_focal = 
		calibration_times %>%
		filter( clade %in% focal_calibrations_clades ) %>%
		arrange( age )

	noised_dates = 
		replicate( 
			noised_replicate_n, 
			add_noise_calibration_times( calibration_times_focal, sd_fraction ),
			simplify = FALSE
		)

	nodes_contrast_noised_list = lapply(
		noised_dates,
		function( x ){
			gene_trees_pruned %>% 
				calibrate_trees( x, model="correlated" ) %>% 
				add_pics_to_trees() %>% 
				summarize_contrasts()
		}
	)

	p_noised = lapply( nodes_contrast_noised_list, wilcox_oc )

@@

This manuscript was computed on `r format( Sys.time(), "%a %b %d %X %Y" )` with the following R package versions.

```{r session_summary, echo=FALSE, comment=NA}
	sessionInfo()
```

```{r wrap_up, echo=FALSE}
	save.image()
```

>>>>

```{r load_data, echo=FALSE, message=F, warning=F}

	# Load data and analysis results generated by 
	# manuscript_kernel.R
	load("manuscript.RData")

@@

@@

@@

@@

@@

@@

```{r file_status, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
	commit_render = system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")

All files needed to re-execute the analyses presented in this document are available at https://github.com/caseywdunn/comparative_expression_2017. The most recent commit at the time the analyses were executed was `r commit_kernel %>% strtrim(8)`. The most recent commit at the time of the manuscript was rendered was `r commit_render %>% strtrim(8)`.

@@

@@

@@

@@

@@

The analyses presented in this manuscript were computed on `r format( system_time_kernel, "%a %b %d %X %Y" )` with the following R package versions.

```{r session_summary, echo=FALSE, comment=NA}
	session_info_kernel
```

This manuscript was rendered from these analyses on `r format( system_time_kernel, "%a %b %d %X %Y" )`.
